---
title: "Using subgroupTree for Subgroup Detection"
author: "Matt Olson and Richard Berk"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Background ##

  In randomized, controlled treatment (RCT) studies, some subsets of the population may benefit from an intervention more than others. Some subsets may also be harmed. One can define these outcomes as a local average treatment effect in which the average treatment effect for the subset differs substantially from the average treatment effect for the all study subjects. For example, in a clinical study measuring the effect of a drug on a certain disease, it may be the drug may be especially effective for women of a certain age. Alternatively, such women may experience harmful side effects.

The aim of the `subgroupTree` package is to identify subsets of a study subjects that display particularly large positive or large negative average treatment effects. The procedure searches for such subgroups though a greedy tree growing procedure. At each stage, the procedure recursively seeks the single predictor and split leading to the largest observed average treatment effect between the subsets defined by that split.

It is also desirable to conduct statistical tests for any local average treatment effect, typically for a null hypothesis that the average local treatment effect equals 0.0. In this setting, classical inferential procedures break down because the process requires adaptive search and selection. We implement a permutation procedure to provide conservative inference for any detected effects. Currently, this is implemented only for subgroups defined by a split on a single predictor based on a tree stump. Larger trees present computational challenges. Details can be found in Berk et al., *Using Recursive Partitioning to Find and Estimate Heterogenous Treatment Effects In Randomized Clinical Trials*  Journal of Experimental Criminology, forthcoming, 2020.

## Simulated Example ##

We will illustrate these ideas through a simple simulation.  The data frame
below contains four predictors `z1`, `x1`, `z2`, `x2`, a treatment indicator
variable `treated`, and a binary response `response` that is conditioned on each
of these variables.  By construction, the the expected values of `response` is
0.9 for treated individuals for which `z1 %in% c('A', 'B') & x1 > 0.5`, and
is 0.5 otherwise.  In other words, this data set contains a subset that
displays an average treatment effect of 0.4, and we hope to detect this subset
with our procedure.

```{r message=FALSE, warning=FALSE}
set.seed(1234)
n = 2000
df = data.frame(z1 = sample(c('A', 'B', 'C', 'D'), n, replace = TRUE),
                x1 = runif(n),
                z2 = sample(c('a', 'b', 'c', 'd'), n, replace = TRUE),
                x2 = runif(n),
                treated = sample(c(FALSE, TRUE), n, replace = TRUE))
high_ate = df$z1 %in% c('A', 'B') & df$x1 > 0.5
df$response = rbinom(n, 1, prob = ifelse(high_ate & df$treated, 0.9, 0.5))
```

The `subgroup_tree` function implements our greedy search procedure and relies
heavily on the `rpart` package.  In the call below, we specify that we
intend to search for **large** treatment effects by specifying `direction = 'max'`,
and we pass in two parameters that govern the tree-growing process.  Specifically,
we search for trees of depth two, and we consider only splits of nodes that
contain 100 or more observations.

```{r, message=FALSE, warning=FALSE}
library(subgroupTree)
max_tree = subgroup_tree(df$response, df$treated, df[,1:4],
                         direction = 'max', maxdepth = 2, minbucket = 100)
print(max_tree)
```

Each line of the output provides information about the structure of the tree
and the average treatment effect computed at each node. We can see from line 7
that the tree detects an average treatment effect of 0.422 among units for
which `z1 == B & x1>=0.5326`. Alternatively, it can be
more convenient to view the same output in graphical form.

```{r, fig.show='hold', fig.align='center', message=FALSE, warning=FALSE}
library(rpart.plot)
rpart.plot(max_tree, extra=1)
```

For illustration, we also demonstrate the effect of changing `direction = 'min'`.
In this case, we are directing the tree to search for subgroups with small
treatment effects (i.e. negative, but large in absolute value).

```{r, message = FALSE, warning = FALSE}
min_tree = subgroup_tree(df$response, df$treated, X = df[,1:4],
                         direction = 'min', maxdepth = 1, minbucket = 100)
print(min_tree)
```

In the tree above, we detect a treatment effect among individuals for which
`x1 < 0.055`.  Of course, by construction the population treatment effects
in this group is zero.  We illustrate our inferential procedure to confirm this.
Each iteration of our permutation procedure shuffles the response and calculates
the largest possible treatment effect t-value that can result splitting the data
into two groups.  At the moment, our implementation only handles single splits
of the data for computational tractability.

```{r, message=FALSE, warning=FALSE}

# sampled t-values
perm_t_score = subgroup_perm_test(df$response, df$treated, df[,1:4],
                                  direction = 'min', ate = 0.0, minbucket = 100)

# t-value for different in
subgroup = (df$x1 < 0.055)
t_score = with(df, t.test(response[treated & subgroup],
                          response[!treated & subgroup], alternative = 'less'))

# permutation p-value
mean(perm_t_score <=  t_score$statistic)
```






